---
title: "KAMP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{KAMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Hello and welcome to the `KAMP` package! 
This package is designed to calculate the expectation and variance of KAMP (K adjustment by Analytical Moments of the Permutation distribution) for point patterns with marks. The package is partially built on the `spatstat` package, which is a powerful tool for analyzing spatial data in R.
The `KAMP` package provides functions to simulate point patterns, calculate the KAMP CSR, and visualize the results. The package is designed to be user-friendly and easy to use, with a focus on providing clear and concise output.
The package is still in development, and we welcome any feedback or suggestions for improvement. If you have any questions or issues, please feel free to reach out to us.

# Before You Begin

A few concepts that come up throughout this vignette:

- **`ppp` object**: `spatstat`'s representation of a point pattern -- a set of point locations (`x`/`y` coordinates) inside a bounding window, optionally with a `marks` attribute labeling each point (e.g. by cell type). `kamp()` can build one for you from a plain data.frame, or you can pass one in directly.

- **Ripley's K function**: A measure of spatial clustering. For a given radius `r`, K(r) roughly captures the expected number of additional points within distance `r` of a typical point, normalized by point density. Values higher than expected under randomness indicate clustering; lower values indicate dispersion (points spread out more than expected).

- **Complete spatial randomness (CSR)**: The null hypothesis that points are distributed independently and uniformly at random. The *theoretical* K under CSR (`theo_csr`) assumes a homogeneous point density across the whole window -- an assumption that's often violated in real tissue images (e.g. due to holes, folds, or regions with no cells).

- **KAMP**: Rather than comparing observed K only to the theoretical CSR baseline, KAMP computes the expectation and variance of K under a *permutation* null -- what you'd see if the marks were randomly reassigned among the existing point locations. This adapts to the actual inhomogeneity of the tissue. KAMP computes this permutation distribution's moments analytically (in closed form) instead of by simulating many permutations, so it stays fast even on large datasets.

# Setup
```{r}
library(KAMP)
#library(devtools)
library(tidyverse)
library(spatstat.random)
#devtools::load_all()
set.seed(50)
```

# Quick Start

Here's the whole workflow in a few lines, before we walk through it in more detail below: load the data, subset to a single image, and compute the KAMP expectation.

```{r}
data(ovarian_df)
one_sample <- ovarian_df %>% filter(sample_id == unique(sample_id)[1])

quick_kamp <- kamp(df = one_sample,
                   rvals = seq(0, 100, by = 10),
                   univariate = TRUE,
                   mark_var = "immune",
                   mark1 = "immune")
quick_kamp
```

`k` is the observed K for the "immune" cells at each radius `r`, and `kamp_csr` is KAMP's adjusted null expectation for that same radius. The rest of this vignette walks through what these numbers mean, how to visualize them, and how to run bivariate and large-scale (KAMP-lite) analyses.

# Ovarian Dataset

The `ovarian_df` dataset is a small dataframe that contains a snapshot of 5 images of ovarian cancer cells from the `HumanOvarianCancerVP()` dataset in the `VectraPolarisData` package. Each image is represented by a unique sample ID, and within each image, there are multiple cells with their respective x and y coordinates. The dataset includes an `immune` column that indicates whether the cell is an immune cell or a background cell. There is also a `phenotype` column that indicates the type of cell, with levels "helper t cell", "cytotoxic t cell", "b cell", "macrophage", "tumor", and "other". The `x` and `y` columns represent the coordinates of the cells in the image.

```{r}
data(ovarian_df)
head(ovarian_df)
```

Since we have a dataframe of multiple images, let's go through, subset our dataframe by id, and plot it.
```{r}
ids <- unique(ovarian_df$sample_id)
mark_var <- "immune"

for (id in ids) {
  df_sub <- ovarian_df %>% filter(sample_id == id)
  w <- convexhull.xy(df_sub$x, df_sub$y)
  pp_obj <- ppp(df_sub$x, df_sub$y, window = w, marks = df_sub[[mark_var]])
  
  p <- ggplot(as_tibble(pp_obj), aes(x, y, color = marks)) +
    geom_point(size = 0.6) +
    labs(title = paste("Sample:", id)) +
    theme_minimal()
  
  print(p)
}
```

# KAMP

Now that we have our data, we can use the `kamp()` function to calculate the KAMP expectation and variance for both univariate and bivariate data.

## Univariate

We can use the `kamp()` function to calculate the KAMP expectation for univariate data. 

The `kamp()` function has several parameters that allow us to customize the calculation:

- `df`: Either a point pattern object created using the `ppp()` function from the `spatstat` package, or a plain data.frame with `x`/`y` columns and a marks column (in which case `mark_var` is required).

- `rvals`: A sequence of distances at which to calculate the K function.

- `univariate`: A logical value indicating whether to calculate the univariate K function (default is `TRUE`).

- `mark_var`: The name of the marks column in `df`, when `df` is a data.frame. Ignored when `df` is already a `ppp` object.

- `mark1`: The value of the marks variable for the first mark (required for univariate).

- `mark2`: The value of the marks variable for the second mark (optional for univariate).

- `variance`: A logical value indicating whether to calculate the variance (default is `FALSE`).

- `correction`: The edge correction to use -- `"trans"`/`"translational"` (the default), `"iso"`/`"isotropic"`, or `"none"`.

- `thin`: A logical value indicating whether to use thinning (default is `FALSE`).

- `p_thin`: Percentage to thin by (default is `0.5`).

For univariate data, we only need to specify one marks variable.
In this case, we set `univariate = TRUE` and `variance = FALSE` (the default).

- `univariate = TRUE` calculates the K function for one mark versus background.

- `variance = FALSE` means we only compute the expectation.

- `correction` uses translational correction by default.

**Choosing a correction:** edge correction matters because points near the window's boundary have fewer visible neighbors than points in the interior, which can bias K downward at larger radii if left uncorrected.

- `"trans"`/`"translational"` (the default): a good default for most cases; reweights each pair of points by how much the window overlaps with itself when shifted by their separation vector.
- `"iso"`/`"isotropic"`: another commonly-used correction; can behave differently for irregularly-shaped windows and is more computationally expensive.
- `"none"`: no correction at all -- fastest option, but biased whenever edge effects aren't negligible (e.g. large `r` relative to the window's size, or a small/irregular window).

### Subsetting Data

To perform univariate analysis, we can subset our `ovarian_df` dataframe to include only the first image ID and the `immune` marks variable.

```{r}
ids <- unique(ovarian_df$sample_id)
univ_data <- ovarian_df %>% filter(sample_id == ids[1])
mark_var <- "immune"
head(univ_data)
```

### Expectation

We can now calculate the KAMP expectation for the univariate data using the `kamp()` function.
What we are doing here is calculating the KAMP expectation for the `immune` marks variable against the background cells, using a sequence of distances from 0 to 100 with a step of 10. Furthermore, we use the option `univariate = TRUE` and specify the marks variable for the first mark as `mark1 = "immune"` - we are not setting `mark2` since we are only interested in the univariate case.

What's returned is a data frame with the following columns:

- `r`: The distance at which the K function is calculated.

- `k`: The K function value.

- `theo_csr`: The theoretical CSR (Complete Spatial Randomness) value.

- `kamp_csr`: The KAMP CSR value.

- `kamp`: The difference between K and the KAMP CSR.


```{r}
univ_kamp <- kamp(univ_data, 
                  rvals = seq(0, 100, by = 10),
                  univariate = TRUE,
                  mark_var = mark_var,
                  mark1 = "immune")

univ_kamp
```


We can visualize KAMP using `ggplot2`. Plotted here is the original K from translational edge correctionm, the theoretical CSR, and the KAMP CSR. The KAMP CSR is the expected value of K under the null hypothesis of CSR, which is calculated using the analytical moments of the permutation distribution.


```{r}
univ_kamp %>%
  ggplot(aes(x = r)) +
  geom_line(aes(y = theo_csr, color = "theo_csr", linetype = "theo_csr"), linewidth = 1) +
  geom_line(aes(y = kamp_csr, color = "kamp_csr", linetype = "kamp_csr"), linewidth = 1) +
  geom_line(aes(y = k, color = "k", linetype = "k"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(
    values = c(
      "theo_csr" = "black",
      "kamp_csr" = "blue",
      "k" = "red"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "theo_csr" = "solid",
      "kamp_csr" = "solid",
      "k" = "dotted"
    )
  ) +
  labs(
    title = "Univariate KAMP Expectation",
    x = "r",
    y = "Value",
    color = "Series",
    linetype = "Series"
  ) +
  theme_minimal()

```


Looking at the plot, we can see that the KAMP CSR (blue line) is slightly higher than the theoretical CSR (black line) at larger distances. This result aligns with our expectations, as, if we view the plots of the first sample image, the first image is more inhomogenous - i.e. there are large holes/"patches" of empty
space that could make the immune cells appear more clustered than expected under CSR.

Let's plot the differences between k and the theo_csr and kamp_csr to get a better idea:

```{r}
univ_kamp %>%
  ggplot(aes(x = r)) +
  geom_line(aes(y = k - kamp_csr, color = "k - kamp_csr", linetype = "k - kamp_csr"), linewidth = 1) +
  geom_line(aes(y = k - theo_csr, color = "k - theo_csr", linetype = "k - theo_csr"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(
    values = c(
      "k - theo_csr" = "red",
      "k - kamp_csr" = "blue"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "k - theo_csr" = "solid",
      "k - kamp_csr" = "solid"
    )
  ) +
  labs(
    title = "Univariate KAMP Expectation - differences",
    x = "r",
    y = "Value",
    color = "Series",
    linetype = "Series"
  ) +
  theme_minimal()
```

As expected, the difference between k and the theoretical CSR values tended to be higher than the difference between k and the KAMP CSR values, indicating that the KAMP CSR seems to do a better job accounting for the inhomogenous tissue quality.

###  Variance

To calculate the variance of the KAMP expectation, we still use the kamp() function, but we set `variance = TRUE`. This will compute the variance of the KAMP expectation at each distance `r`. The new output will include all the same columns as before, but with additional column `var` that contains the variance of the KAMP expectation at each distance, and a `p-val` column that contains the p-value for the variance test.


**Note:** `variance = TRUE` is not compatible with `thin = TRUE`. If both are set to TRUE, a warning will be thrown and variance will still be calculated, but the results may be unreliable.

```{r}
univ_kamp_var <- kamp(univ_data,
                      rvals = seq(0, 100, by = 10),
                      univariate = TRUE,
                      mark_var = mark_var,
                      mark1 = "immune",
                      variance = TRUE)
univ_kamp_var
```

We can visualize the variance of the KAMP expectation using `ggplot2`:

```{r}
#devtools::load_all()

univ_kamp_var %>%
  ggplot(aes(x = r, y = var)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Univariate KAMP Variance", x = "r", y = "Variance") +
  theme_minimal()
```

## Bivariate

### Subsetting Data
For bivariate analysis, we can subset our `ovarian_df` dataframe to include two types of immune cells and background cells. In this example, we will use "helper t cells" and "cytotoxic t cells".

```{r}
ids <- unique(ovarian_df$sample_id)
biv_data <- ovarian_df %>% 
  filter(sample_id == ids[1]) #%>%
#filter(phenotype %in% c("helper t cells", "cytotoxic t cells", "other")) %>%
#droplevels()

mark_var <- "phenotype"

head(biv_data)
```



### Expectation

To calculate the KAMP expectation for bivariate data, we set `univariate = FALSE` and specify the marks variables for both marks.


```{r}

biv_kamp <- kamp(df = biv_data,
                 rvals = seq(0, 100, by = 10),
                 univariate = FALSE,
                 mark_var = mark_var,
                 mark1 = "helper t cell",
                 mark2 = "cytotoxic t cell")
head(biv_kamp)
```

```{r}
biv_kamp %>%
  ggplot(aes(x = r)) +
  geom_line(aes(y = theo_csr, color = "theo_csr", linetype = "theo_csr"), linewidth = 1) +
  geom_line(aes(y = kamp, color = "kamp", linetype = "kamp"), linewidth = 1) +
  geom_line(aes(y = k, color = "k", linetype = "k"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(
    values = c(
      "theo_csr" = "black",
      "kamp" = "blue",
      "k" = "red"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "theo_csr" = "solid",
      "kamp" = "dotted",
      "k" = "dotted"
    )
  ) +
  labs(
    title = "Bivariate KAMP Expectation",
    x = "r",
    y = "Value",
    color = "Series",
    linetype = "Series"
  ) +
  theme_minimal()
```

### Variance

```{r}
biv_kamp_var <- kamp(df = biv_data,
                     rvals = seq(0, 100, by = 10),
                     univariate = FALSE,
                     mark_var = mark_var,
                     mark1 = "helper t cell",
                     mark2 = "cytotoxic t cell",
                     variance = TRUE)
head(biv_kamp_var)
```

```{r}
biv_kamp_var %>%
  ggplot(aes(x = r, y = var)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Bivariate KAMP Variance", x = "r", y = "Variance") +
  theme_minimal()
```

# KAMP-lite (Thinning)

KAMP-lite refers to running KAMP with a thinned point pattern using `thin = TRUE` and specifying `p_thin` between 0 and 1. This helps with performance on large datasets.

## Univariate

### Expectation
```{r}
ids <- unique(ovarian_df$sample_id)
mark_var <- "immune"
univ_data <- ovarian_df %>% filter(sample_id == ids[1])

univ_kamp_lite <- kamp(df = univ_data,
                       rvals = seq(0, 100, by = 10),
                       univariate = TRUE,
                       mark_var = mark_var,
                       mark1 = "immune",
                       thin = TRUE,
                       p_thin = 0.3)
univ_kamp_lite
```

```{r}
univ_kamp_lite %>%
  ggplot(aes(x = r)) +
  geom_line(aes(y = theo_csr, color = "theo_csr", linetype = "theo_csr"), linewidth = 1) +
  geom_line(aes(y = kamp, color = "kamp", linetype = "kamp"), linewidth = 1) +
  geom_line(aes(y = k, color = "k", linetype = "k"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(
    values = c(
      "theo_csr" = "black",
      "kamp" = "blue",
      "k" = "red"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "theo_csr" = "solid",
      "kamp" = "dotted",
      "k" = "dotted"
    )
  ) +
  labs(
    title = "Univariate KAMP-lite Expectation",
    x = "r",
    y = "Value",
    color = "Series",
    linetype = "Series"
  ) +
  theme_minimal()
```

### Variance
```{r}
univ_kamp_lite_var <- kamp(df = univ_data,
                           rvals = seq(0, 100, by = 10),
                           univariate = TRUE,
                           mark_var = mark_var,
                           mark1 = "immune",
                           thin = TRUE,
                           p_thin = 0.3,
                           variance = TRUE) # should display a warning message
univ_kamp_lite_var
```

## Bivariate
```{r}
ids <- unique(ovarian_df$sample_id)
biv_data <- ovarian_df %>% 
  filter(sample_id == ids[1]) #%>%
#filter(phenotype %in% c("helper t cell", "cytotoxic t cell", "other")) %>%
#droplevels()

mark_var <- "phenotype"

head(biv_data)
```

### Expectation
```{r}
biv_kamp_lite <- kamp(df = biv_data,
                      rvals = seq(0, 100, by = 10),
                      univariate = FALSE,
                      mark_var = mark_var,
                      mark1 = "helper t cell",
                      mark2 = "cytotoxic t cell",
                      thin = TRUE,
                      p_thin = 0.3)
head(biv_kamp_lite)
```

```{r}
biv_kamp_lite %>%
  ggplot(aes(x = r)) +
  geom_line(aes(y = theo_csr, color = "theo_csr", linetype = "theo_csr"), linewidth = 1) +
  geom_line(aes(y = kamp, color = "kamp", linetype = "kamp"), linewidth = 1) +
  geom_line(aes(y = k, color = "k", linetype = "k"), linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_color_manual(
    values = c(
      "theo_csr" = "black",
      "kamp" = "blue",
      "k" = "red"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "theo_csr" = "solid",
      "kamp" = "dotted",
      "k" = "dotted"
    )
  ) +
  labs(
    title = "Bivariate KAMP-lite Expectation",
    x = "r",
    y = "Value",
    color = "Series",
    linetype = "Series"
  ) +
  theme_minimal()
```

### Variance
```{r}
biv_kamp_lite_var <- kamp(df = biv_data,
                          rvals = seq(0, 100, by = 10),
                          univariate = FALSE,
                          mark_var = mark_var,
                          mark1 = "helper t cell",
                          mark2 = "cytotoxic t cell",
                          variance = TRUE,
                          thin = TRUE,
                          p_thin = 0.3) # should display a warning message
head(biv_kamp_lite_var)
```

```{r}
biv_kamp_lite_var %>%
  ggplot(aes(x = r, y = var)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Bivariate KAMP lite Variance", x = "r", y = "Variance") +
  theme_minimal()
```

# Troubleshooting

A few messages/errors you may encounter, and what they mean:

- **"We expect the dataframe to be a single point process..."** -- this is an informational message (not an error), printed every time `kamp()`/`check_inputs()` runs. If your dataframe actually contains multiple images/samples, subset it to one `sample_id` (as we do throughout this vignette) before calling `kamp()`, and loop over samples yourself.

- **`mark1 is not a mark in the point pattern object`** -- the value you passed to `mark1` (or `mark2`) doesn't match any level of the marks column exactly, including case and whitespace. Check `unique(your_df[[mark_var]])` to see the exact values available.

- **`mark_var must be supplied and cannot be NULL or empty`** -- you passed a plain data.frame as `df` but didn't specify `mark_var`. This is only required when `df` isn't already a `ppp` object.

- **`The mark_var column must have at least two unique values`** -- `kamp()` needs at least two distinct mark values (e.g. your mark of interest plus "everything else") to compare against.

- **"Variance calculation with KAMP-lite is not recommended"** -- thinning (`thin = TRUE`) randomly drops points, which adds noise on top of the variance KAMP already estimates; results with both `thin = TRUE` and `variance = TRUE` are usable but should be interpreted cautiously.

# Next Steps

- See the [README](https://dliao1.github.io/KAMP/) for a condensed summary of `kamp()`'s arguments and output columns.
- `?kamp`, `?kamp_expectation`, `?kamp_variance`, `?kamp_expectation_biv`, and `?kamp_variance_biv` document every argument in detail, including the lower-level functions that `kamp()` dispatches to.
- Questions or issues? Please open one on [GitHub](https://github.com/dliao1/KAMP/issues).

```{r}
sessionInfo()
```

