---
title: "Introduction to kamp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to kamp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
Hello and welcome to the `kamp` package! 
This package is designed to calculate the expectation and variance of KAMP (K adjustment by Analytical Moments of the Permutation distribution) for point patterns with marks. The package is partially built on the `spatstat` package, which is a powerful tool for analyzing spatial data in R.
The `kamp` package provides functions to simulate point patterns, calculate the KAMP CSR, and visualize the results. The package is designed to be user-friendly and easy to use, with a focus on providing clear and concise output.
The package is still in development, and we welcome any feedback or suggestions for improvement. If you have any questions or issues, please feel free to reach out to us.

# Setup
```{r}
library(kamp)
#library(devtools)
library(tidyverse)
library(spatstat.random)
#devtools::load_all()
set.seed(50)
```

# Simulating Data
The `sim_pp_data` function can be used to simulate univariate point patterns, while the `sim_pp_data_biv` function can be used to simulate bivariate point patterns.

The `sim_pp_data` function takes the following arguments:

- `lambda_n`: The number of points to simulate.

- `abundance`: The abundance of the point pattern.

- `markvar1`: The name of the first cell type. Defaults to "immune".

- `markvar2`: The name of the second cell type. Defaults to "background".

- `distribution`: The distribution of the point pattern. This can be either "hom" for homogeneous or "inhom" for inhomogeneous.

- `clust`: A logical value indicating whether to cluster the points or not.


The `sim_pp_data_biv` function takes the same arguments as `sim_pp_data`, but also includes an additional argument for the third cell type.

- `markvar3`: The name of the third cell type. Defaults to "background".

The `sim_pp_data` and `sim_pp_data_biv` functions return a `ppp` object, which is a class in the `spatstat` package that represents a point pattern. The `ppp` object contains the x and y coordinates of the points, as well as the marks associated with each point.

## Univariate

```{r}
univ_data <- sim_pp_data(lambda_n = 200, 
                    abundance = 0.3,
                    markvar1 = "immune",
                    markvar2 = "background",
                    distribution = "hom",
                    clust = FALSE)

```

We can plot this using `ggplot`:

```{r}
as_tibble(univ_data) %>%
    ggplot(aes(x,y, color = marks)) +
    geom_point()
```

## Bivariate
```{r}
biv_data <- sim_pp_data_biv(lambda_n = 200, 
                    abundance = 0.3,
                    markvar1 = "immune1",
                    markvar2 = "immune2",
                    markvar3 = "background",
                    distribution = "inhom",
                    clust = TRUE)

```

We can plot this using `ggplot`:

```{r}
as_tibble(biv_data) %>%
    ggplot(aes(x,y, color = marks)) +
    geom_point()
```

# Expectation

The `kamp_expectation` function calculates the expectation of KAMP for a given point pattern. The function takes the following arguments:

- `ppp`: The point pattern object. 

- `rvec`: A vector of distances at which to calculate the KAMP expectation. Defaults to c(0, 0.05, 0.075, 0.1, 0.15, 0.2)

- `correction`: Edge correction method. This can either be "iso" or "trans" for isotropic or translation correction, respectively. The default is "trans".

- `markvar`: Identifies subset of marked points. Defaults to immune.

- `thin_pct`: The percentage of points to thin from the point pattern. This is used to reduce the number of points in the pattern and speed up the calculation. The default is 0.

The `kamp_expectation` function returns a data frame with the following columns:

- `r`: The distance at which the KAMP expectation was calculated.

- `K`: The KAMP expectation value.

- `theo_csr`: The theoretical CSR value.

- `kamp_csr`: The KAMP CSR value.

- `kamp_fundiff`: The difference between K and the KAMP CSR.

Running this code below calculates the KAMP expectation for the univariate point pattern.
```{r}
univ_kamp_expectation <- kamp_expectation(ppp_obj = univ_data,
                                          rvec = c(0, 0.05, 0.1, 0.15, 0.2),
                                          correction = "trans",
                                          markvar = "immune",
                                          thin_pct = 0)

univ_kamp_expectation
```

Here is the matrix-based equivalent:
```{r}
univ_kamp_expectation_mat <- kamp_expectation_mat(ppp_obj = univ_data,
                                                  rvec = c(0, 0.05, 0.1, 0.15, 0.2),
                                                  correction = "trans",
                                                  markvar = "immune",
                                                  thin_pct = 0)

univ_kamp_expectation_mat
```

We can plot the `kamp_fundiff` values for the univariate point pattern using `ggplot`:
```{r}
as_tibble(univ_kamp_expectation) %>%
    ggplot(aes(r, kamp_fundiff)) +
    geom_point() +
    geom_line() +
    labs(x = "Distance (r)", y = "fundiff = K - KAMP CSR") +
    theme_minimal()
```

Likewise, this code calculates the KAMP expectation for the bivariate point pattern.
```{r}
biv_kamp_expectation <- kamp_expectation_biv(ppp_obj = biv_data,
                                             rvec = c(0, 0.05, 0.1, 0.15, 0.2),
                                             correction = "trans",
                                             markvar1 = "immune1",
                                             markvar2 = "immune2",
                                             thin_pct = 0
                                             )
biv_kamp_expectation
```

And here is the matrix-based equivalent:
```{r}
biv_kamp_expectation_mat <- kamp_expectation_biv_mat(ppp_obj = biv_data,
                                                     rvec = c(0, 0.05, 0.1, 0.15, 0.2),
                                                     correction = "trans",
                                                     markvar1 = "immune1",
                                                     markvar2 = "immune2",
                                                     thin_pct = 0
                                                     )
biv_kamp_expectation_mat
```

We can plot the `kamp_fundiff` values for the bivariate point pattern using `ggplot`:
```{r}
as_tibble(biv_kamp_expectation) %>%
    ggplot(aes(r, kamp_fundiff)) +
    geom_point() +
    geom_line() +
    labs(x = "Distance (r)", y = "fundiff = K - KAMP CSR") +
    theme_minimal()
```


# Variance

The `kamp_variance` function calculates the variance of KAMP for a given point pattern. The function takes the following arguments:

- `ppp`: The point pattern object.

- `rvec`: A vector of distances at which to calculate the KAMP variance. Defaults to c(0, 0.05, 0.075, 0.1, 0.15, 0.2)

- `correction`: Edge correction method. This can either be "iso" or "trans" for isotropic or translation correction, respectively. The default is "trans".

- `markvar`: Identifies subset of marked points. Defaults to immune.

- `thin_pct`: The percentage of points to thin from the point pattern. This is used to reduce the number of points in the pattern and speed up the calculation. The default is 0.


The `kamp_variance` function returns a data frame with the following columns:

- `r`: The distance at which the KAMP variance was calculated.

- `K`: The KAMP variance value.

- `theo_csr`: The theoretical CSR value.

- `kamp_csr`: The KAMP CSR value.

- `kamp_fundiff`: The difference between K and the KAMP CSR.

- `var`: The variance of the KAMP expectation.

- `z`: The z-score of the KAMP expectation.

- `pval`: The p-value of the KAMP expectation.

Note: there is no spatstat equivalent for the variance function.

Running this code below calculates the KAMP variance for the univariate point pattern.
```{r}
univ_kamp_variance <- kamp_variance(ppp_obj = univ_data,
                                          rvec = c(0, 0.05, 0.1, 0.15, 0.2),
                                          correction = "trans",
                                          markvar = "immune",
                                          thin_pct = 0)

univ_kamp_variance
```

Here it is for the bivariate point pattern:
```{r}
biv_kamp_variance <- kamp_variance_biv(ppp_obj = biv_data,
                                       rvec = c(0, 0.05, 0.1, 0.15, 0.2),
                                       correction = "trans",
                                       markvar1 = "immune1",
                                       markvar2 = "immune2",
                                       thin_pct = 0
                                       )

biv_kamp_variance
```

# Ovarian Dataset
The `ovarian_df` dataset is a point pattern dataset that contains the x and y coordinates of ovarian cancer cells, as well as the sample ID and the mark variable. The mark variable indicates whether the cell is an immune cell or a background cell.

The `ovarian_df` dataset is a small dataframe that contains a snapshot of 5 images of ovarian cancer cells from the `HumanOvarianCancerVP()` dataset in the `VectraPolarisData` package. 

. Each image is represented by a unique sample ID, and the x and y coordinates of the cells are provided in the `x` and `y` columns, respectively. The `marks` column contains the mark variable, which indicates whether the cell is an immune cell or a background cell.


```{r}
data(ovarian_df)
head(ovarian_df)

```

```{r}
ids = unique(ovarian_df$sample_id)

ovarian_df <- ovarian_df %>% filter(sample_id == ids[[1]])


marksvar = "immune"
w = convexhull.xy(ovarian_df[["x"]], ovarian_df[["y"]])
pp_obj_ovarian = ppp(ovarian_df[["x"]], ovarian_df[["y"]], window = w, marks = ovarian_df[[marksvar]])

as_tibble(pp_obj_ovarian) %>%
    ggplot(aes(x,y, color = marks)) +
    geom_point() +
    labs(title = "Ovarian Cancer Cells") +
    theme_minimal()
```


