---
title: "my-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(kperm)
library(tidyverse)
library(spatstat.random)

```

```{r}
sim_data <- get_sim_data(lambda_n = 200, 
                    abundance = 0.3,
                    type = "hom",
                    clust = FALSE)
marksvar = "immune"
rvalues = seq(0, 1, length.out = 50)

#w = convexhull.xy(ovarian[["x"]], ovarian[["y"]])

# define pp_obj
# pp_obj = ppp(ovarian[["x"]], ovarian[["y"]], window = w, marks = ovarian[[marksvar]])

pp_obj <- sim_data

df_back <- as.data.frame(pp_obj) %>%
  as_tibble()

df_back


as_tibble(pp_obj) %>%
    ggplot(aes(x,y, color = marks)) +
    geom_point()
```

move helpers to utils file

no need for timing stuff
think abt user facing stuff
10000 cells use border correction
thinning percentage = 0
whole separate function for bivariate
unit tests with small spatstat generated datasets
https://github.com/julia-wrobel/registr/blob/master/R/register_fpca.R (documentation example)

SIMULATION SETUP:
can compare traditional K (trans border correction), with Kinhom, Kperm (with 100 permutations), and KAMP


images should be: random, clustered, and inhomogenous (generate and show images)

4 nulls total:
2 null: NO CLUSTERING and CLUSTERING
within each, have homogenous and inhomogenous background


Varied rate of total cells (n) and immune cell abundance (p)
n ∈(100, 500, 1000) # maybe tweak bc i decreased
p ∈(0.01, 0.1, 0.5)

3*3 = 9 images for each subset null hypothesis so 9x4 = 36 total, assuming im avging for the barplots and stuff?

Compare: accuracy (bar plots of fundiff and should be centered at 0) and computation time

can also compare power and type I error # yah idk how i can compare ? if fundiff > 0 thats clustering? are we getting the variance for the perm and KAMP dist and comparing CIs?? need to look at the code



